---
title: "Reproducible Research: Assessment 1"
author: "Sean McHugh"
date: "17 June 2020"
output: 
  html_document:
    keep_md: true
---

## <b>Introduction: Why are we here?</b>
This is the first assignment for the Reproducible Research Course which is part of the Johns Hopkins University Data Science Specialization hosted on Coursera. The purpose of this assignment will be to demonstrate the ability to utilize R markdown files and the knitr process to transform the explanation and display of the R code for the exploration of data into a 'human readable' format. To accomplish this goal, the assignment utilizes activity monitoring data collected on personal digital devices. The dataset we will explore is a CSV file with a total of 17,568 observations (according to the assignment prompt) of activity step counts per 5 minute intervals, from October and November 2012.


## <b>Prompt 0 - Loading and preprocessing the data</b>
In order to prepare the data, we will preload the R packages for use in this project, download the data from the URL provided in the assignment prompt, unzip the data file, and read the file into memory.

#### Step 0.1 - Load the R Packages Required for this Assignment
```{r,echo=TRUE,message=FALSE}
library(dplyr)
library(ggplot2)
library(data.table)
```

#### Step 0.2 - Create Source Data Variables
```{r,echo=TRUE}
SourceURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
SourceFileName <- "repdata_data_activity.zip"
SourceDest <- "activity.csv"
```

#### Step 0.3 - Verify/Download the Source Data
```{r,echo=TRUE}
if(!file.exists(SourceFileName)) {
  download.file(SourceURL,SourceFileName,method='curl')
}
```

#### Step 0.4 - Verify/Unzip the Source Data
```{r,echo=TRUE}
if(!file.exists(SourceDest)) {
  unzip(SourceFileName)
}
```

#### Step 0.5 - Read the Source Data
```{r,echo=TRUE}
activityData <- read.csv("activity.csv")

```
#### Step 0.6 - Format the Dates into date format
```{r,echo=TRUE}
activityData$date <- as.Date(activityData$date,format="%Y-%m-%d")
```

#### Step 0.7 - Confirm the structure of the data
```{r, echo=TRUE}
head(activityData)
summary(activityData)
```

## <b>Prompt 1 - What is mean total number of steps taken per day?</b>

### Prompt 1-1: Calculate the total number of steps taken per day
First, we need to create a new variable 'dailyTotalSteps' in which to store the summary table of daily total steps.
```{r,echo=TRUE}
dailyTotalSteps <- activityData %>% group_by(date) %>% summarise(dailySteps=sum(steps))
```
Next, we want to display the head of the new variable to verify we achieved the correct transformation.
```{r,echo=TRUE}
head(dailyTotalSteps)
```

### Prompt 1-2: If you do not understand the difference between a histogram and a barplot, research the difference between them. Make a histogram of the total number of steps taken each day
```{r,echo=TRUE}
hist(dailyTotalSteps$dailySteps,main = "Histogram: Daily Total Steps Measured October-November 2012",xlab = "Total Daily Steps",ylab="Frequency (# of Days)")
```

### Prompt 1-3: Calculate and report the mean and median of the total number of steps taken per day
For an easy calculation of the mean and median of total daily steps taken, we can use the summary function:
```{r,echo=TRUE}
summary(dailyTotalSteps)
```
### <b style="color:#117A65"><i>Answer: Mean daily steps is 10766. Median daily steps is 10765.</i></b>

## <b>Prompt 2 - What is the average daily activity pattern?</b>

### Prompt 2-1: Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
First we need to create a new table that rather than summarizing by date, instead summarizes by 5-minute interval.
```{r,echo=TRUE}
dailyTotalAVG <- activityData %>% filter(!is.na(steps)) %>% group_by(interval) %>% summarise(dailyAVGSteps=mean(steps))

head(dailyTotalAVG)
dailyTotalAVG
plot(dailyTotalAVG$interval,dailyTotalAVG$dailyAVGSteps,type="l",main="Average Steps per Daily 5-Minute Interval (from OCT-NOV 2012)", ylab="# of Steps", xlab="5-Minute Intervals over 24 Hours")

```

### Prompt 2-2: Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
We should be able to answer this by calculating the maximum value directly utilizing the 'which.max' function:
```{r,echo=TRUE}
maxSteps <- dailyTotalAVG[which.max(dailyTotalAVG$dailyAVGSteps),][[1]]
maxSteps
```
#### <b style="color:#117A65"> The 5-Minute Interval `r maxSteps` on average contains the most steps from the entire dataset , which covers the time period from 08:30-08:35</b>
## <b>Prompt 3 - Inputing missing values</b>
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

### Prompt 3-1: Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

### Prompt 3-2: Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

### Prompt 3-3: Create a new dataset that is equal to the original dataset but with the missing data filled in.

### Prompt 3-4: Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?


## <b>Prompt 4 - Are there differences in activity patterns between weekdays and weekends?</b>
For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

### Prompt 4-1: Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
    
### Prompt 4-2: Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.